---
title: "Structural Equation Model"
author: "Scott Large"
date: "6/12/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# install.packages("lavaan")
# remotes::install_github("rstudio/nomnoml")
# remotes::install_github("thomasp85/tidygraph")

library(lavaan)
library(nomnoml)
library(tidygraph)
library(ggraph)
```


## Introduction


## Materials and Methods

# Data

```{r}
sim_dat <- read.csv("~/projects/soe-synthesis/analysis/data/simulated-dataset/simulated_biomass_catch.csv")
sim_temp <- read.csv("~/projects/soe-synthesis/analysis/data/simulated-dataset/simulated_temperature.csv")

```


```{r}


bio <- sim_dat %>% 
  select(Year, Species, Biomass) %>%
  group_by(Year) %>%
  dplyr::mutate(Biomass = scale(Biomass, center = TRUE, scale = TRUE)) %>% 
  tidyr::pivot_wider(names_from = Species, values_from = Biomass) %>% 
  as.data.frame()

catch <- sim_dat %>% 
  select(Year, Species, catch) %>%
  group_by(Year) %>%
  dplyr::mutate(Species = paste0(Species, "_c"), 
                catch = scale(catch, center = TRUE, scale = TRUE)) %>% 
  tidyr::pivot_wider(names_from = Species, values_from = catch) %>% 
  as.data.frame()

all_dat <- cbind(bio, 
                 catch,
                 temp = sim_temp$Anomaly)


model <- '
  # measurement model
    system_biomass =~ Herring + Cod + Yellowtail + Goosefish
  # regressions
    system_biomass ~ temp
  # residual correlations
    Herring ~~ Cod
    Herring ~~ Herring
'

fit <- sem(model, data=all_dat)
summary(fit, standardized=TRUE)
plot(fit)
sp_coef


```



```{r}
sp_coef <- rbind(
  c("Atlantic_herring",	0.00000001, 1.40E-07, 0.000000605, 0),
  c("Atlantic_cod",	0, 4.20E-07, 0.0000008, 0),
  c("Goosefish", 0, 2.97E-07, 1.00E-06, 0),
  c("Yellowtail",	0, 0, 0, 7.50E-07))

colnames(sp_coef) <- c("to", "Atlantic_herring", "Atlantic_cod", "Goosefish", "Yellowtail")

sp_graph <- tibble::as_tibble(sp_coef) %>% 
  tidyr::pivot_longer(-to, names_to = "from", values_to = "weight") %>% 
  select(from,
         to,
         weight) %>% 
  as_tbl_graph()


create_layout(sp_graph, layout = "auto")

sp_graph %>% 
  ggraph(layout = "hive") + 
  geom_edge_link() + 
  geom_node_point(size = 8, colour = 'steelblue') +
  geom_node_text(aes(label = name), colour = 'black', vjust = 0.4) + 
  ggtitle('Joining graphs') + 
  theme_graph()

```

# Modeling
We used SEM to further understand the direct and inderect effects of different environmental and biological variables. SEM is a statistical approach to parameterize and test hypothesized relationships, described as causal models between multiple variables, to solve a set of equations implied by a path diagram (@ref/Hox @ref/Grace). We compared maximum likelihood SEM with covariance matricies to explore linear and additive relationships and XYZ with abc to explore nonlinear relationships ([nlsem](https://www.jstatsoft.org/article/view/v077i07), Umbach et al 2017). . The SEM provides a more powerful way of identifying relationships among variables that are hierarchcially dependent (Stewart 2014)

Structural equation models were fit using the `lavaan` package (Rosseel, 2011) in R by bootstrapping (x draws).



Grace JB, Bollen KA (2005) Interpreting the results from multiple regression and structural equation models. Bulletin of the Ecological Society of America, 86, 283–295.

Hox J, Bechger T (1998) An introduction to structural equation modelling. Family Science Review, 11, 354–373.
